Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-31 paper.tex
reg update
MAJOR REVISION UNDERWAY

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is necessary to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks, including an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency. We explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicate that the VSI
  requires a short thermal relaxation timscale. For vertically
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature (which gives rise to vertical
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. 
\end{abstract}

\section{Introduction}
% stuff about hydrodynamic angular momentum transport 
Understanding the mechanisms of angular mometum transport is the
central problem in the theory of accretion disks
\citep{lydenbell74}. The
magneto-rotational instability in the context of astrophysical disks 
\citep{balbus91} is now accepted as a robust pathway to
magneto-hydrodynamic (MHD) turbulence, which transports angular momentum
outwards and thus enables mass accretion.

>>>>

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations. These simulations indicate that the VSI 
  requires a short thermal relaxation timscale. For vertically 
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. 
\end{abstract}

\section{Introduction}
% stuff about hydrodynamic angular momentum transport 
Understanding the mechanisms of angular mometum transport is the
central problem in the theory of accretion disks
\citep{lyndenbell74}. The  
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

However, MHD turbulence may not be applicable for all 
accretion disks. For example, there may be regions in protoplanetary
disks --- called dead zones ---  which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent state-of-the-art

%in this respect purely hydrodynamical turbulence should be

### 2015-01-31 setup.tex
MAJOR REVISION UNDERWAY

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant, and $H_\mathrm{iso}$ is a
characteristic scale-height.  Thus, $\epsilon$ is also the
characteristic disk aspect-ratio.  
% From $\epsilon$ we can define a
% characteristic scale-height  
% \begin{align}
%   \Hiso \equiv \epsilon r.% \quad\quad(\Gamma \equiv 1),   
% \end{align} 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
and we are interested in neutrally or sub-adiabatically stratified
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

@@

@@

In the this shearing box the equilibrium density, pressure and
sound-speed are assumed to depend only on $z$, by evaluating these
quantities at $r=r_0$ in the global disk. The background rotation is
approximated by a linear shear in $x$, but left unaltered in $z$.

The axisymmetric VGSB equations read 
\begin{align}
  & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
  \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
  v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
  2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
  &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
    v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
  \nabla\cdot\bm{v} -
  \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), \label{energy_eq}
\end{align}
where $\bm{v}$ is the velocity field relative to the bulk flow
(i.e. $\bm{v}=\bm{0}$ in equilibrium),  
subscript $0$ indicates evaluation at $r=r_0$. In the energy equation
the ratio of specific heats $\gamma$ is assumed constant, 
and we have included a thermal relaxation term with timescale
$t_c=\beta\Omega_k^{-1}$ with $\beta$  being a dimensionless input
parameter.

>>>>

We examine disks with power-law profiles for the midplane density
$\rho_0(r)\equiv\rho(r,0)$: 
\begin{align}
  \rho_0(r) = \rho_{00}\left(\frac{r}{r_0}\right)^p,
\end{align}
where $\rho_{00}$ is the midplane density at the fiducial radius
$r=r_0$ and $p$ is the power-law index. The equilibrium pressure $P$
and density $\rho$ are related by 
\begin{align}
  P(r,z) &= 
  \frac{c_{s0}^2(r)}{\Gamma}\rho_0(r)^{1-\Gamma}\rho^\Gamma(r,z),\notag\\
  & \equiv K(r)\rho^{\Gamma}(r,z). \label{eqm_press}
\end{align}
where the polytropic index $\Gamma$ is a parameter and 
the midplane sound-speed $c_{s0}(r)$ is taken to be another power-law: 
\begin{align}
  c_{s0}^2(r)=c_{s00}^2\left(\frac{r}{r_0}\right)^q, 
\end{align}
where $c_{s00}\equiv c_{s0}(r_0)$  and $q$ is the power-law index for
the midplane temperature. We define
\begin{align}
  \epsilon(r) \equiv \frac{c_{s0}(r)}{v_k(r)} \equiv
  \frac{H_\mathrm{iso}}{r} 
\end{align}
to parametrize the magnitude of the sound-speed, where
$v_k=r\Omega_k=\sqrt{GM_*/r}$ is the Keplerian azimuthal velocity and
$G$ is the gravitational constant, and $H_\mathrm{iso}$ is a
characteristic scale-height.  Thus, $\epsilon$ is also the
characteristic disk aspect-ratio.  
% From $\epsilon$ we can define a
% characteristic scale-height  
% \begin{align}
%   \Hiso \equiv \epsilon r.% \quad\quad(\Gamma \equiv 1),   
% \end{align} 
The global sound-speed $c_s(r,z)$ is given by
\begin{align}
  c_s^2\equiv \frac{dP}{d\rho} =
  c_{s0}^2(r)\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = \frac{\Gamma P}{\rho}. 
\end{align}
For discussion purposes it is useful to define the vertical buoyancy frequency $N_z$ via
\begin{align}
  N_z^2 = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   
\end{align}
where $\gamma$ is the constant adiabatic index. We are interested in
neutrally or sub-adiabatically stratified 
disks with $\Gamma\leq \gamma$ so that $N_z^2\geq0$.

@@

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p R} \right)&>0.  
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure (assumed to be constant), and $j\equiv r^2\Omega$ is the
specific angular momentum.

For rotationally-supported, thin disks, pressure gradients are small 
compared to rotation, so the first criterion is generally
satisfied. Thus, to assess stability, we evaluate the second criterion
for the above disk models in $z>0$ where $\p_zP<0$, without loss of
generality,   
\begin{align}
  \gamma - \Gamma > \frac{s\epsilon^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
    -\left(\gamma-\Gamma\right)\frac{\p\ln{P}}{\p\ln{r}} \right]. 
\end{align} 
For typical model parameters the right hand side is
$O(\epsilon^2)$. Hence, to have instability in the adiabatic limit 
requires $\gamma < \Gamma + O(\epsilon^2)$. Since we consider
$\epsilon\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations.

@@

In this shearing box the equilibrium density, pressure and
sound-speed are assumed to depend only on $z$, by evaluating these
quantities at $r=r_0$ in the global disk. The background rotation is
approximated by a linear shear in $x$, but left unaltered in
$z$. However, the vertical shear $\p_z\Omega$ is also assumed to only
depend on $z$.

The axisymmetric VGSB equations read 
\begin{align}
  & \frac{\p\rho}{\p t} + \frac{\p}{\p x}\left(\rho v_x\right) +
  \frac{\p}{\p z}\left(\rho v_z\right)=0,\\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
  v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}} = - 
  2\Omega_0\hat{\bm{z}}\times\bm{v} - S_0v_x\hat{\bm{y}} \notag\\ 
  &\phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} +
    v_zr_0\frac{d\Omega_0}{dz} \hat{\bm{y}}=}-\frac{1}{\rho}\nabla P - \frac{\p\Phi_0}{\p z}\hat{\bm{z}},\\
  & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P = -\gamma P
  \nabla\cdot\bm{v} -
  \frac{\rho}{t_c}\left(\frac{P}{\rho}-\left.\frac{P}{\rho}\right|_{t=0}\right), \label{energy_eq}
\end{align}
where $\bm{v}$ is the velocity field relative to the bulk flow
(i.e. $\bm{v}=\bm{0}$ in equilibrium),  
subscript $0$ indicates evaluation at $r=r_0$. In the energy equation 
we have added a thermal relaxation term with timescale
$t_c=\beta\Omega_k^{-1}$ with $\beta$  being a dimensionless input
parameter.

\subsection{Limitations}
Before using the VGSB to study the VSI, it is very important to be
aware of several limitations of the box model. The VGSB momentum  
equation have the same form as that for the standard shearing box 
\citep{goldreich65}, except for the appearence of the term 
$v_zr_0d\Omega_0/dz$. This term appears in the global stability
problem and is the only term associated with vertical shear that is 
ported from the global disk into the local model.

%horizontal structure not ported

Note that, within the VGSB framework, the vertical shear 
$d\Omega_0/dz$ is a free parameter (as are $\Omega_0$ and $S_0$). 
This is different to the global problem because in that case, the
vertical shear is a result of the disk radial structure, which is
ignored in the VGSB equilibria. 
%because there is only one
%equilibrium condition, which determines the vertical density/pressure
%structure. 
In effect, we are studying dynamics in the standard shearing box
subject to an additional azimuthal force $-v_zr_0 d \Omega_0/dz$. This
force is inspired by the vertical shear in the global disk, and its
functional form is taken from global equilibrium. 
%but should be considered as a model parameter

%horizontal structure that gives rises 
%vertical shear is an input force
%only porting vertical shear  
%vertical shear doesn't depend on x 
%not all results can be extrapolated to the global context

In Appendix we make the connection between the linear problem in the
VGSB to that in the global disk model.

### 2015-01-31 summary.tex
MAJOR REVISION UNDERWAY

@@

\emph{Geometry.} Studying the VSI in the shearing box has an 
inconsistency. Namely, the equilibrium temperature in the
box only depends on height (if at all). However, it is precisely the
radial temperature (or entropy) gradient in the global disk that
gives rise to vertical shear. This is not a serious issue for
perturbations with small radial wavelengths, 
%explicitly checked by including global terms 
but will need to be resolved by adopting global disk models  
if one wishes to generalize the VSI to large-scale perturbations.

>>>>

%relation with kley simulations

@@

\emph{Geometry.} Extrapolating our results from the shearing box to a global disk  
has an inconsistency. Namely, the equilibrium temperature in the
box only depends on height (if at all). However, it is precisely the
radial temperature (or entropy) gradient in the global disk that
gives rise to vertical shear. We do not expect this to be a serious issue for 
perturbations with small radial wavelengths, 
%explicitly checked by including global terms 
but will need to be resolved by adopting global disk models  
if one wishes to generalize the VSI to large-scale perturbations.
